---
title: 'Week 5: R Functions'
author: "Thrisha Praveen"
date: "2025-01-27"
output: pdf_document
---

This week, we're introducing **R functions** and how to write our own functions.

Questions from lab document:

> Q1. Write a function grade() to determine an overall grade from a vector of student homework
assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an
NA value) this can be used as a score to be potentially dropped. Your final function should be
adquately explained with code comments and be able to work on an example class gradebook
such as this one in CSV format: “https://tinyurl.com/gradeinput” [3pts]

```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)

student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)

student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)

```

Follow the guidelines from class:
- Write a working snippet of the code that solves a simple problem

```{r}
#straight forward mean()
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
mean(student1)
```

...But, we need to drop lowest grade

```{r}
#which element of the vector is the lowest
which.min(student1)
```

What I want is to now drop (i.e exclude) this lowest score from the mean() calculation
```{r}
#returns everything but the 8th element in vector
student1[-8]

```

Now, we can use the answer from which.min() to return all other elements of the vector.
```{r}
#first working snippet
mean(student1[-which.min(student1)])

```

What about other example students? Will this work for them?

We could try using na.rm=TRUE argument, but it's not a good approach because it's unfair.
```{r}
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
mean(student2, na.rm=TRUE)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
mean(student3, na.rm=TRUE)
```

Another approach: mask/replace all NA's with an 0. How do we find the NA values?
```{r}
x <- student2

is.na(x)
which(is.na(x))
```

Now that we have identified the NA elements, we need to mask/replace them. Replace them with 0?

```{r}
#code that's useful for masking NA
x[is.na(x)] <- 0
x

mean(x)
```

Recall, we should drop the lowest score.

```{r}
x[is.na(x)] <- 0
mean(x[-which.min(x)])

```

Now, we are essentially there with our working snippet.
```{r}
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
x <- student3
x[is.na(x)] <- 0
mean(x[-which.min(x)])
```

##Now, we make our function!

Take the snippet & turn it into a function. Every function has 3 parts:
- name, in our case `grade()`
- input args, a vector of student scores
- the body, our working snippet of code

Using RStudio, I can select `code > extract function`
```{r}
grade <- function(x) {
  x[is.na(x)] <- 0
  mean(x[-which.min(x)])
}
grade(student1)
grade(student2)
grade(student3)
```

We now need comments to explain the code to our future selves and others who want to understand the code.

```{r}
#' Calculate average score for a vector of scores, dropping the lowest score. 
#' Missing values are treated as zeroes.
#'
#' @param x A numeric vector of scores
#'
#' @returns Average score
#' @export
#'
#' @examples
#' student <- c(100, NA, 90, 97)
#' grade(student)
#' 
grade <- function(x) {
  #mask NA with 0
  #treat missing values as 0
  x[is.na(x)] <- 0
  #exclude lowest score from mean
  mean(x[-which.min(x)])
}
```

Now, we can use our function on our real, whole class data from this:
CSV format: “https://tinyurl.com/gradeinput”

```{r}
url <- "https://tinyurl.com/gradeinput"
gradebook <- read.csv(url, row.names =1)
```

```{r}
apply(gradebook, 1, grade)
```

> Q2. Using your grade() function and the supplied gradebook, Who is the top scoring student
overall in the gradebook? [3pts]

To answer this, we run the apply() function and save the results.
```{r}
results <- apply(gradebook, 1, grade)
sort(results, decreasing=TRUE)
```
```{r}
which.max(results)
```

> Q3. From your analysis of the gradebook, which homework was toughest on students (i.e. obtained
the lowest scores overall? [2pts]

```{r}
gradebook
```

```{r}
ave.scores <- apply(gradebook, 2, mean, na.rm=TRUE)
ave.scores
which.min(ave.scores)
```
```{r}
med.scores <- apply(gradebook, 2, median, na.rm=TRUE)
med.scores
which.min(med.scores)
```
```{r}
boxplot(gradebook)
```
The boxplot reveals that HW2 was the toughest on students and had the lowest score overall.

>Q4. Optional Extension: From your analysis of the gradebook, which homework was most
predictive of overall score (i.e. highest correlation with average grade score)? [1pt]

Are the final results (avg score per student) correlated with the scores for individual homeworks (gradebook columns).
```{r}
masked.gradebook <- gradebook
masked.gradebook[is.na(masked.gradebook)] <-0
masked.gradebook
#cor(results, masked.gradebook$hw1)
#cor(results, masked.gradebook$hw5)

```
And look at correlation
```{r}
cor(results, masked.gradebook$hw5)

apply(masked.gradebook, 2, cor, x=results)
```
HW5 has the best correlation between the grade for a specific assignment and overall grade/result in the class.

>Q5. Make sure you save your Quarto document and can click the “Render” (or Rmarkdown”Knit”) button to generate a PDF foramt report without errors. Finally, submit your PDF

Knit the document to make a PDF
to gradescope. [1pt]
