---
title: "class10"
format: pdf
prefer-html: true
---

1.  Importing candy data

```{r}
candy_file <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv")

candy = data.frame(candy_file, row.names=1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
ncol(candy)
```

But 3 of those columns aren't candy types, so there are 9 total candy types.

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

2.  What is your favorite candy?

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["100 Grand", ]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent

```

Side-note: the skimr::skim() function

```{r}
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The winpercent variable/column seems to be on a different scale than the other variables/columns of the dataset since its mean is 50.316...%, while the other columns' means are between 0-1.

> Q7. What do you think a zero and one represent for the candy\$chocolate column?

It represents whether a specific candy contain chocolate. The candy would get a 0 in the candy\$chocolate column if it doesn't contain chocolate and get a 1 if it does.

3.  Overall Candy Rankings \>Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], n=5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(candy$winpercent, decreasing=TRUE), ], n=5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()+
  theme(
    axis.text.y = element_text(size = 4)  # Reduce text size
  )
```

Adding useful color to the bar chart:

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) +
  theme(
    axis.text.y = element_text(size = 4)  # Reduce text size
  )
```

Now, for the first time, using this plot we can answer questions like: \> Q17. What is the worst ranked chocolate candy?

Sixlets

> Q18. What is the best ranked fruity candy?

Nik L Nip

4.  Taking a look at pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures are ranked the highest in terms of winpercent for the least money (located in the bottom right quadrant-ish of the plot).

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

Nik L Nip is the most expensive and the least popular. Here are the top 5 most expensive candies:

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

```{r}
# Make a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

5.  Exploring the correlation structure

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

chocolate & fruity, bar & pluribus

> Q23. Similarly, what two variables are most positively correlated?

winpercent & chocolate, also bar and chocolate

6.  Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

Now we can plot our main PCA score plot of PC1 vs PC2.

```{r}
plot(pca$x[,1], pca$x[,2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)

```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)

```

```{r}
ggplotly(p)

```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, and pluribus. This makes sense since a lot of hard fruit candies are sold together in big packs, like Life Savers.
